\Sof{TBA - New Titles in the comments}

### 2020-05-04 paper/icsme20.tex
new version

% \title{Fixing Vulnerabilities Hinders Maintainability}
\title{\Sof{TBA - New Titles in the comments}}
%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
%\title{The reason why you should be applying coding practices: safer software.}
%\title{What is the impact of security patches in software evolution?}
%\title{Hey, developers! You should pay more attention to your patches.}

@@

From a methodological 
perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 
applied in the code bases to patch vulnerabilities affect code maintainability. Especially, for patches that address \emph{Improper Input Validation} (CWE-20), \emph{Information 
Exposure} (CWE-200), \emph{Improper Restriction of Operations within the Bounds of a 
Memory Buffer} (CWE-119), \emph{Missing Release of Memory after Effective Lifetime} 
(CWE-401) and \emph{Path Traversal} (CWE-22). This is important work
because there is still little information about the impact of security patches on the software
maintainability. With this study, 
we intend to highlight the need for tools to predict the impact of patches on 
maintainability~\cite{4724577}, documentation to help developers adopt security 
best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
security-related code evolution in computer curricula.
%
%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
%

@@

\textit{\textbf{RQ1: What is the impact of security patches on the
maintainability of open-source software?}}
\textit{Patching vulnerabilities may harm the maintainability of open-source
software.}

In this section, we report and discuss the impact of patches on open-source
software maintainability under four different views: guidelines/metrics, 
severity, programming language and formula proposed - the mean of all 
guidelines/metrics.

% As expected, vulnerabilities with higher severity are more likely to impact negatively
% the maintainability of software. This suggests that high/medium severity vulnerabilities may
% need more attention than low severity when patching vulnerabilities.
% %
% We would expect that the negative impact of programming languages on
% maintainability to be more severe, as arguably poor design of programming
% languages and lack of using the best practices lead to more buggy and vulnerable
% code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
% Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
% the result of projects contributions policies. Nine of the top-10 of dataset projects
% with more contributors have restrict contribution policies regarding code
% standards. Some examples are the \texttt{laravel/framework},
% \texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
% Standard}\footnote{PSR-2 Standard available at
% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.

@@

Although overall patching vulnerabilities has a less negative impact, it is 
important to note that the design debt of one guideline can lead
to severe impacts on the software quality~\cite{10.1145/1985362.1985366} 
and in the market value of a company~\cite{4267025}. Thus,
tools like Better Code Hub and static analysis should be leveraged to 
assess the risk of patching vulnerabilities - alongside with pull
requests and code reviews. While style-guided static code analysis
may be daunting due to the amount of rules and different effects
on maintainability, BCH ensures that the $10$ chosen guidelines are
the ones with the highest effect on maintainability. 
%
% 
 \begin{figure}[H]
  	\centering
 	\vspace{-0.3 cm}
  	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
 	\vspace{-0.3 cm}
  	\caption{Impact of the security patches per guideline.}
 	\label{fig:guidelines}
 	\vspace{0.2 cm}
 	\scalebox{0.5}{
 	\begin{tabular}{r@{: }l r@{: }l}
 	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
 	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns 
	in Modules \\
 	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
 	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of 
	Code
 	\end{tabular}
 	}	
 \end{figure}

\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
Database} (NVD) website to obtain their severity levels. In total, we 
retrieved severity scores for $551$ vulnerabilities: $117$ (\emph{HIGH}), 
$404$ (\emph{MEDIUM}) and $30$ (\emph{LOW}). Figure~\ref{fig:severity} 
presents the impact of security patches per severity level on the 
maintainability of open-source software. We observe that patches for 
\emph{HIGH} ($47.9\%$) and \emph{MEDIUM} ($44.1\%$) severity vulnerabilities 
hinder more the maintainability of software than \emph{LOW} ($30.0\%$) 
severity vulnerabilities. Again, patches have a considerable 
negative impact on software maintainability - between $30\%$ 
and $50\%$. None statistical significance was retrieved. However, 
results should not be disregarded because they seem to confirm the 
assumption that higher severity vulnerabilities patches may have a 
more negative impact on maintainability.

@@

\textbf{Programming Languages:} We have also evaluated the impact on the maintainability per programming 
language of open-source software (Figure~\ref{fig:lang_main}). We restrict 
this analysis to programming languages with at least $20$ data points, as this 
is a requirement for our hypothesis tests. Thus, we compare the results for 
\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
\texttt{Python} and \texttt{PHP}. The remaining languages are included into 
\texttt{Other}. \texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and 
\texttt{PHP} ($38.27\%$) are the programming languages with a more negative
impact on the maintainability of software. \texttt{Java} seems to be the 
language less affected by patching, more than $50\%$ of patches had a positive 
impact on maintainbility. But overall patches in all languages have a 
considerable negative impact on maintainability - from $30\%$ to $50\%$ - which 
confirms the need for better and more secure programming languages. 
Statistical significance was only retrieved for the \texttt{C/C++} 
(p-value = $2.24$x$10^{-05}$) and \texttt{Ruby} (p-value = $0.041$) languages.
Results should be read with care where statistical significance was not found.
Yet, data reports very interesting hints on the impact of programming languages 
on security patches.

@@

@@

\textbf{Formula:} Maintainability is being calculated as the mean of 
the guidelines computed by Better Code Hub (BCH). Regarding security 
commits results (Figure~\ref{fig:secvsreg}), we observe that
maintainability decreases in $41.28\%$ ($412$), remains the same in $20.64\%$ 
($206$) and increases in $38.08\%$ ($380$) of the cases. The resulting
p-value of the Paired Wilcoxon signed-rank test is $0.044$. Since the p-value is 
below the significance level of $0.05$, we argue that \emph{security patches 
have a negative impact on the maintainability of open-source software}.
%

@@

\emph{Weakness}, according to Common Weakness Enumeration (CWE) glossary, 
is a type of mistake that, in proper conditions, could contribute to the 
introduction of vulnerabilities within that product. This enumeration
system was used to classify the vulnerabilities included in our study.
Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
the impact of all patches grouped using the first-level weaknesses from
the \emph{Research Concepts} list on CWE; and, the remaining charts
show two of those first-level weaknesses more in depth reporting the impact 
on maintainability of their sub-weaknesses (\emph{(b) and (c)}).

@@

\emph{Cross-Site Scripting (CWE-79)} is the weakness that suffers less 
positive/negative impact from security patches, i.e., patching usually 
has no impact on the maintainability of open-source software. This can 
be explained by the fact that patches to this weakness usually are easily 
done by adding an \texttt{escape} function. Patches of \emph{SQL Injection
(CWE-89)} weaknesses equally hinder and improve the software maintainability. 
But overall a very considerable percentage of patches hinder the open-source 
software maintainability - between $30\%$ and $60\%$.

Statistical significance was only found to the CWE-119 weakness (p-value = 
$0.041$). However, results obtained to the other weaknesses should not be 
disregarded because they offer an interesting view on how the maintainability 
is impacted per weakness which can provide insights to developers/researchers
on the weaknesses that need more attention.

\textit{\textbf{RQ3: What is the impact of security patches versus regular 
patches on the maintainability of open-source software?}}

The impact of security and baseline changes in the maintainability of
open-source software is presented in Figure~\ref{fig:secvsreg}.  
We have seen, previously, a deterioration on software maintainability 
when patching vulnerabilities: $41.28\%$ ($412$) suffered a 
negative impact, $20.64\%$ ($206$) remained the same and $38.08\%$ 
($380$) increased the software maintainability. For regular changes, 
we observe that the maintainability decreases in $27.05\%$ ($270$) 
and increases in $30.26\%$ ($302$) of the cases. But in contrast to 
security patches, the maintainability of regular changes remains the 
same in $42.69\%$ ($426$) of the cases and performing regular
changes has a more positive impact than negative on the software 
maintainbility.

Results show that regular changes typically have no impact on the 
software maintainability which contradicts previous research.
However, we manually inspected a considerable amount of cases from
the regular changes distribution where this happens and found that 
identifying regular commits as similar in size as the security-related 
commit is limiting the type of regular commits (e.g, input 
refactorings, variables or functions, type conversion, etc).  
One example is the regular commit\footnote{Regular change: 
\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad32
2448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated 
by our methodology and using as reference the following security commit
\footnote{Security change: 
\url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc
445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
overflow vulnerability patch, our methodology chose a similiar
type of change. We strongly believe that this phenomenon lead
that increased significantly the number of cases with no impact
on the software maintainability.

The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
conclusions regarding the regular changes are not statistically significant. 
Yet, it is important to note that \emph{patching security vulnerabilities seem 
to have a more negative impact on the maintainability of open-source software 
sthan performing regular changes}.

@@

% \section{Discussion}\label{sec:discussion}
% In this section, we discuss the results and answer the proposed research
% questions.
% \textit{\textbf{RQ1: What is the impact of security patches on the
% maintainability of open-source software?}}
% \textit{Patching vulnerabilities may harm the maintainability of open-source
% software.}
% The guideline analysis suggests that all guidelines are affected by security patches. Being the
% \emph{Write Short Units of Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns in Modules} the guidelines with higher negative impact on the maintainability ($>30\%$ of cases). \emph{Keep Architecture Components Balanced} is the guideline with lower negative impact on
% the maintainability (approx. $10\%$). This might imply that developers should give different
% levels of attention to the guidelines when patching vulnerabilities.
% %
% As expected, vulnerabilities with higher severity are more likely to impact negatively
% the maintainability of software. This suggests that high/medium severity vulnerabilities may
% need more attention than low severity when patching vulnerabilities.
% %
% We would expect that the negative impact of programming languages on
% maintainability to be more severe, as arguably poor design of programming
% languages and lack of using the best practices lead to more buggy and vulnerable
% code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
% Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
% the result of projects contributions policies. Nine of the top-10 of dataset projects
% with more contributors have restrict contribution policies regarding code
% standards. Some examples are the \texttt{laravel/framework},
% \texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
% Standard}\footnote{PSR-2 Standard available at
% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
% \subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
% affect open-source software maintainability?}}}
% \textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
% open-source software maintainability.}
% Overall, the results obtained for the maintainability per CWE are no statistical significance
% The impact of a patch depends on its complexity, i.e., if the patchs
% adds complexity to the code base it is probably affecting the software
% maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} (CWE-79)
% patches endure more
% cases without any impact in the maintainability. Typically, these types of
% vulnerabilities do not need extra lines to be fixed, as shown in
% Listing~\ref{lst:fix}. Whereas, \emph{Resource Management Errors} (CWE-399) and vulnerabilities
% such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
% attacks are more difficult to fix (several code lines deleted and added).
% Another example is the Apple authentication flaw discovered in $2017$
% (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
% \url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
% could log in as root with an empty password. Patrick Wardle examined the cause
% of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
% \url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
% concluded that the flaw was due to the introduction of high cyclomatic complexity
% in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
% developers' preoccupations along with the other patterns that explicitly hinder
% maintainability because despite their low complexity these are patterns that
% still produce a considerable negative impact on the maintainability. Real
% efforts have been made in creating and incorporating coding standards and best
% practices in software production. However, security is still far from perfection
% and fixing vulnerabilities might hinder software maintainability. Thus, it is
% not only important to provide tool support to developers to help them apply these
% patterns without harming software maintainability, but also to create and redesign
% programming languages to facilitate developers improving software maintainability.
% We measured the effect sizes between both distribution -- before and after
% patches --- per language and security patterns using the Common Language Effect
% Size~\cite{cliff:1993} method. Effect sizes are of a low-medium level.
% Thus, considering the $0$ median and the low effect sizes, it is reasonable
% to affirm that there is a significant effect but it might only be
% recognized after a thoughtful data analysis.
% \subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes
% on the maintainability of open-source software?}}}
% \textit{Patching security vulnerabilities has a larger negative and positive impact
% than when fixing regular commits.}
% We found that $41.76\%$ of the security patches have a negative impact on the
% maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). While $21.04\%$ of regular commits have a negative impact on the maintainability. This raises
% a new trade-off when developers need to patch vulnerabilities on their projects
% because they might be affecting the software maintainability or even introducing
% new vulnerabilities. This study exhibits evidence that developers may have to
% reduce maintainability for the sake of security. We understand that developers
% should be able to produce secure code without affecting the maintainability of
% their projects. But there are still some concerns that need to be addressed:
% \begin{itemize}
% 	\item \textbf{Documentation} of software must provide developers with the
% 	best practices to implement secure and maintainable software. Unmaintainable
% 	code is difficult to test, analyze and re-use.
% 	\item\textbf{Programming Languages} should provide new design patterns to
% 	easily implement security patterns without endangering maintainability.
% 	Ultimately, new programming languages, both secure and maintainable by design,
% 	may be designed to help developers be highly productive when writing and
% 	maintaining  secure applications. One example is the need for designing new
% 	authorization mechanisms since these are one of the most complex security
% 	features to implement.
% 	\item \textbf{Software Developers} lack the knowledge to create secure and
% 	maintainable code. Mainly because academic curricula are not yet prepared
% 	to properly educate them. Universities should do their part on making
% 	this possible. Online services such as BCH have also a great potential to help
% 	developers to be more aware of the maintainability issues introduced by their
% 	changes. Thus, developers can put more work into improving their code maintainability
% 	and avoid common issues such as code duplication.
% \end{itemize}

\section{Threats to Validity}\label{sec:threats}
%
The following section presents the potential threats to the validity of this
study.
%
\\\textit{\textbf{Construct:}} The maintainability formula was inferred based 
on the BCH's reports. The high amount of different projects and backgrounds 
may require other maintainability standards. However, BCH does use a 
representative benchmark of closed and open-source software projects to compute 
the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
%
\\\textit{\textbf{Internal:}} The security patches dataset provided by previous
work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
commits produced by project developers to classify the changes performed by them
to fix security flaws. This approach discards security patches that were
not explicit in commits messages. Pontas et. al\cite{10.1109/MSR.2019.00064} dataset
has patches that were manually mined in software repositories. Being a manual technique,
the search might led to
wrong associations between vulnerabilities and code snippets.
Baseline commits are retrieved randomly by selecting one commit from the same
project of a given security patch. This approach softens the differences
that may result from the characteristics of each project. However,
maintainability may still be affected by the developers' experience, coding
style and software contribution policies which is not evaluated in this study.
Furthermore, this evaluation considers that $998$ regular commits are enough to
alleviate random irregularities in the maintainability differences of the
baseline.
%
\\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.
However, the methodology requires to access data that is not publicly available.
Thus, our findings may not extend to non-open source software. Different programming 
languages may require different coding practices to address software safety. The 
dataset comprises more commits in Java, i.e., the dataset may not be representative 
of the population regarding programming languages. Our approach does not consider 
security patches in any other languages but English.

>>>>

% \title{Fixing Vulnerabilities Hinders Maintainability}
\title{\Sof{TBA - New Titles in the comments}}
%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
%\title{The reason why you should be applying coding practices: safer software.}
%\title{What is the impact of security patches in software evolution?}
%\title{Hey, developers! You should pay more attention to your patches.}
%\title{Are good coding pratices used when patching software vulnerabilities?}

@@

From a methodological 
perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 
applied in the code bases to patch vulnerabilities affect code maintainability. Especially, 
for patches that address \emph{Improper Restriction of Operations within the Bounds of 
a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200).
This is important work
because there is still little information about the impact of security patches on the software
maintainability. With this study, 
we intend to highlight the need for tools to predict the impact of patches on 
maintainability~\cite{4724577}, documentation to help developers adopt security 
best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
security-related code evolution in computer curricula.
%
%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
%

@@

\textit{\textbf{RQ1: What is the impact of security patches on the
maintainability of open-source software?}}

In RQ1, we report and discuss the impact of patches on open-source
software maintainability under four different views: guidelines/metrics, 
severity, programming language and formula proposed - the mean of all 
guidelines/metrics.

@@

Although overall patching vulnerabilities has a less negative impact, it is 
important to note that the design debt of one guideline can lead
to severe impacts on the software quality~\cite{10.1145/1985362.1985366} 
and in the market value of a company~\cite{4267025}. Thus,
tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
ESLint, Infer, and more) should be leveraged to 
assess the risk of patching vulnerabilities - alongside with pull
requests and code reviews. While static code analysis
may be daunting due to the number of rules and different effects
on maintainability, BCH only uses the top-$10$ of the guidelines
with the highest effect on maintainability. Static analysis 
can be used to complement the source code analysis, introduce
the capability of vulnerability detection, and, consequently,
reduce the risk of hindering maintainability.

Nevertheless, results show that developers may fail to follow the coding
practice guidelines and hinder software maintainability after patching; and, 
that different levels of attention should be paid to each guideline. For instance,
\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
are the most affected guidelines.

%
% 
 \begin{figure}[h]
  	\centering
 	\vspace{-0.3 cm}
  	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
 	\vspace{-0.3 cm}
  	\caption{Impact of the security patches per guideline.}
 	\label{fig:guidelines}
 	\vspace{0.2 cm}
 	\scalebox{0.5}{
 	\begin{tabular}{r@{: }l r@{: }l}
 	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
 	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns 
	in Modules \\
 	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
 	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of 
	Code
 	\end{tabular}
 	}	
 \end{figure}

\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
Database} (NVD) website to obtain their severity levels. In total, we 
retrieved severity scores for $551$ vulnerabilities: $117$ (\emph{HIGH}), 
$404$ (\emph{MEDIUM}) and $30$ (\emph{LOW}). Figure~\ref{fig:severity} 
presents the impact of security patches per severity level on the 
maintainability of open-source software. We observe that patches for 
\emph{HIGH} ($47.9\%$) and \emph{MEDIUM} ($44.1\%$) severity vulnerabilities 
hinder more the maintainability of software than \emph{LOW} ($30.0\%$) 
severity vulnerabilities. Again, patches have a considerable 
negative impact on software maintainability - between $30\%$ 
and $50\%$. None statistical significance was retrieved. However, 
results should not be disregarded because they seem to confirm the 
assumption that higher severity vulnerabilities patches may have a 
more negative impact on maintainability. This suggests that high/medium 
severity vulnerabilities may need more attention than low severity while 
patching vulnerabilities which was also observed in previous work but 
regarding the companies market values~\cite{4267025}.

@@

\textbf{Programming Languages:} We have also evaluated the impact on the maintainability per programming 
language of open-source software (Figure~\ref{fig:lang_main}). We restrict 
this analysis to programming languages with at least $20$ data points, as this 
is a requirement for our hypothesis tests. Thus, we compare the results for 
\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
\texttt{Python} and \texttt{PHP}. The remaining languages are included into 
\texttt{Other}. \texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and 
\texttt{PHP} ($38.27\%$) are the programming languages with worse 
impact on the maintainability of software. \texttt{Java} seems to be the 
language less affected by patching, more than $50\%$ of patches had a positive 
impact on maintainbility. But overall patches in all languages have a 
considerable negative impact on maintainability - from $30\%$ to $50\%$ - which 
confirms the need for better and more secure programming languages. 
Statistical significance was only retrieved for the \texttt{C/C++} 
(p-value = $2.24$x$10^{-05}$) and \texttt{Ruby} (p-value = $0.041$) languages.
Results should be read with care where statistical significance was not found.
Yet, data reports very interesting hints on the impact of programming languages 
on security patches.

@@

We would expect that the negative impact of programming languages on
maintainability to be more severe, as arguably poor design of programming
languages and lack of using the best practices lead to more buggy and vulnerable
code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
a significant negative impact approximate to $50\%$ on software 
maintainability. We suspect that these values are
the result of projects contributions policies. Nine of the top-10 of dataset projects
with more contributors have restrict contribution policies regarding code
standards. Some examples are the \texttt{laravel/framework},
\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
instance, \texttt{laravel/framework} requires the \texttt{PSR-2
Standard}\footnote{PSR-2 Standard available at
\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.

@@

\textbf{Overall Score:} Maintainability is being calculated as the mean of 
the guidelines computed by Better Code Hub (BCH). Regarding security 
commits results (Figure~\ref{fig:secvsreg}), we observe that
maintainability decreases in $41.28\%$ ($412$), remains the same in $20.64\%$ 
($206$) and increases in $38.08\%$ ($380$) of the cases. The resulting
p-value of the Paired Wilcoxon signed-rank test is $0.044$. Since the p-value is 
below the significance level of $0.05$, we argue that \emph{security patches 
may have a negative impact on the maintainability of open-source software}; and, 
that tools such as Better Code Hub should be integrated in the CI/CD pipeline
to help developers evaluate the risk of patches hindering software maintainability
alongside with Pull Requests/Code Reviews.
%

@@

\emph{Weakness}, according to Common Weakness Enumeration (CWE) glossary, 
is a type of mistake that, in proper conditions, could contribute to the 
introduction of vulnerabilities within that product. The CWE enumeration
system was used to classify the vulnerabilities included in our study.
Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
the impact of all patches grouped using the first-level weaknesses from
the \emph{Research Concepts} list on CWE; and, the remaining charts
present the impact on maintainability of the sub-weaknesses of two of
the first-level weaknesses (\emph{(b) and (c)}).

@@

The impact of a patch depends on its complexity, i.e., if the patche
adds complexity to the code base it is probably affecting the software
maintainability. \emph{Cross-Site Scripting (CWE-79)} patches endure more
cases with no impact on the open-source software maintainability.
Typically, these types of vulnerabilities do not need extra lines 
to be fixed, as shown in Listing~\ref{lst:fix} - one simple \texttt{escape} 
function patches the issue.  Whereas, \emph{Missing Release of Memory after Effective 
Lifetime (CWE-401)} and \emph{Resource Management Errors} (CWE-399) vulnerabilities
are usually the root cause of Denial-of-Service attacks which are more difficult
to patch since it usually requires to add more complexity to the program - as
seen in the example presented on Section~\ref{sec:motivation}.

Patches of \emph{SQL Injection
(CWE-89)} weaknesses equally hinder and improve the software maintainability. 
But overall a very considerable percentage of patches hinder the open-source 
software maintainability - between $30\%$ and $60\%$.

\textit{Improper Restriction of Operations within the Bounds of 
a Memory Buffer (CWE-119), Path Traversal (CWE-22), Missing Release of Memory after Effective 
Lifetime (CWE-401) and \emph{Information Exposure 
(CWE-200)} are more likely to affect open-source software maintainability.}
Statistical significance was only found to the CWE-119 weakness (p-value = 
$0.041$). However, results obtained to the other weaknesses should not be 
disregarded because they offer an interesting view on how the maintainability 
is impacted per weakness which can provide insights to developers/researchers
on the weaknesses that need more attention.

% Another example is the Apple authentication flaw discovered in $2017$
% (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
% \url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
% could log in as root with an empty password. Patrick Wardle examined the cause
% of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
% \url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
% concluded that the flaw was due to the introduction of high cyclomatic complexity
% in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
% developers' preoccupations along with the other patterns that explicitly hinder
% maintainability because despite their low complexity these are patterns that
% still produce a considerable negative impact on the maintainability. Real
% efforts have been made in creating and incorporating coding standards and best
% practices in software production. However, security is still far from perfection
% and fixing vulnerabilities might hinder software maintainability. Thus, it is
% not only important to provide tool support to developers to help them apply these
% patterns without harming software maintainability, but also to create and redesign
% programming languages to facilitate developers improving software maintainability.
%

\textit{\textbf{RQ3: What is the impact of security changes versus regular 
changes on the maintainability of open-source software?}}

The impact of security and baseline changes in the maintainability of
open-source software is presented in Figure~\ref{fig:secvsreg}.  
We have seen, previously, a deterioration on software maintainability 
when patching vulnerabilities: $41.28\%$ ($412$) suffered a 
negative impact, $20.64\%$ ($206$) remained the same and $38.08\%$ 
($380$) increased the software maintainability. For regular changes, 
we observe that the maintainability decreases in $27.05\%$ ($270$) 
and increases in $30.26\%$ ($302$) of the cases. But in contrast to 
security patches, the maintainability of regular changes remains the 
same in $42.69\%$ ($426$) of the cases and performing regular
changes has a more positive impact than negative on the software 
maintainability.

Results show that regular changes typically have no impact on the 
software maintainability which contradicts previous research\Sof{missing citation}.
However, we manually inspected a considerable amount of cases from
the regular changes distribution where this happens and found that 
identifying regular commits as similar in size as the security-related 
commit is limiting the type of regular commits (e.g, input 
refactorings, variables or functions, type conversion, etc).  
One example is the regular commit\footnote{Regular change: 
\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad32
2448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated 
by our methodology and using as reference the following security commit
\footnote{Security change: 
\url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc
445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
overflow vulnerability patch, our methodology chose a similiar
type of change. We strongly believe that this phenomenon lead
to the significant number of cases where there is no impact
on the software maintainability.

Results show that for both types of changes there is a negative impact
on software maintainability - especially, for security changes. This raises 
a new trade-off when developers need to patch vulnerabilities on their projects
because they might be hindering the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to patch and produce secure code without hindering the maintainability of
their projects. But there are still some concerns that need to be addressed
and that this study tries to bring awareness for:

\begin{itemize}

	\item\textbf{Computer Curricula Needs to be Updated:}
	Software Developers lack the knowledge to create secure and
	maintainable code. Mainly because academic curricula is not yet prepared
	to properly educate them. Universities should do their part on making
	this possible. Online services such as BCH have also a great potential to help
	developers to be more aware of the maintainability issues introduced by their
	changes - alongside with pull requests and code reviews and integrated in 
	the CI/CD pipeline. Thus, developers can 
	put more work into improving their code maintainability and avoiding the risk
	of introducing new vulnerabilities after patching older ones.

	\item\textbf{Secure Programming Languages Wanted:} Programming Languages 
	should provide new design patterns to easily implement security patterns 
	without endangering maintainability.
	Ultimately, new programming languages, both secure and maintainable by design,
	may be designed to help developers be highly productive when writing and
	maintaining secure applications. One example is the need for designing new
	authorization mechanisms since these are one of the most complex security
	features to implement.

	\item\textbf{Poor Coding Practices Documentation:} 
	Software documentation must provided to developers with the
	best practices to implement secure and maintainable software. Unmaintainable
	code is difficult to test, analyze and re-use. Software companies must
	do an effort to provide these artifacts to their developers, in order,
	to avoid the introduction of new vulnerabilities while patching older ones.

\end{itemize}

The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
conclusions regarding the regular changes are not statistically significant. 
Yet, it is important to note that \emph{patching security vulnerabilities seem 
to have a more negative impact on the maintainability of open-source software 
than performing regular changes}.

@@

\section{Threats to Validity}\label{sec:threats}
%
The following section presents the potential threats to the validity of this
study.
%
\\\textit{\textbf{Construct:}} The maintainability formula was inferred based 
on the BCH's reports. The high amount of different projects and backgrounds 
may require other maintainability standards. However, BCH does use a 
representative benchmark of closed and open-source software projects to compute 
the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
The way how maintainability is being calculated - mean of all guidelines - is
a preliminar measure to show how the maintainability is overall affected. Nevertheless,
in some cases the maintainability for the two versions might be the same but have
different causes, i.e., being affected by different guidelines. 
%
\\\textit{\textbf{Internal:}} The security patches dataset provided by previous
work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
commits produced by project developers to classify the changes performed by them
to fix security flaws. This approach discards security patches that were
not explicit in commits messages. Pontas et. al\cite{10.1109/MSR.2019.00064} dataset
has patches that were manually mined in software repositories. Being a manual technique,
the search might led to wrong associations between vulnerabilities and code snippets. 
However, we manually inspected all the data points again and disregarded the ones we were
in doubt. 
When comparing security changes versus regular changes, our study does not take into consideration that 
developers may patch the issue quickly and later proceed to refactor.
We assume that vulnerabilities are patched once or in different parts consecutively.
Baseline commits are retrieved randomly by selecting one commit with an 
approximate size from the same
project of a given security patch. This approach softens the differences
that may result from the characteristics of each project. However,
maintainability may still be affected by the developers' experience, coding
style and software contribution policies which is not evaluated in this study.
Furthermore, this evaluation considers that $998$ regular commits are enough to
alleviate random irregularities in the maintainability differences of the
baseline. 
%
\\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.
However, the methodology requires to access data that is not publicly available.
Thus, our findings may not extend to non-open source software. Different programming 
languages may require different coding practices to address software safety. The 
dataset comprises more commits in Java, i.e., the dataset may not be representative 
of the population regarding programming languages. Our approach does not consider 
security patches in any other languages but English.
